Relict deciduous oaks show high resilience to drought despite the effect of land-use legacies

### 2019-05-02 man/ms.Rmd
update

During the second half of the twentieth century the Iberian Peninsula suffered several extreme drought episodes (*e.g*. 1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014; @Guerreiro2017].The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2014; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017] and they were characterized as extreme drought in our climatic data (Figures S1-S2; Table S3). We focused on these two drought events because they were those included in the period where there is availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events and analyze forest resilience to drought in a longer term (see Table S3). The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Spinoni2015. We used SPEI-12 months scale for all 0.5º grid cells covering Sierra Nevada obtained from Global SPEI Database (http://spei.csic.es/database.html). A severe drought event starts in the month when SPEI falls below the threshold of -1.28 [@Spinoni2017a; @Pascoa2017]. A drought event is considered only when SPEI values are below the threshold for at least two consecutive months [*e.g.* @Spinoni2015; @Spinoni2017a]. Several indicators were computed for each drought event: ***duration*** as the number of consecutive months with the SPEI lower than a certain threshold; ***severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event; the ***intensity*** and the ***Lowest SPEI*** refer to the mean and lowest value of SPEI respectively during the drought event duration.

@@

The analysis of growth changes revealed differences in forest history between sites (Figures 5 and S4). Northern site (SJ) showed two release events (GC > 50 %) detected at stand-wise scale (occurring in more than 50 % of sampled trees): the first during the 1940 decade and the second in the period 1995-2000. These periods alternate with periods of supression. Southern sites (CA-High and CA-Low) showed no release events except for CA-High at the beginning of the 1830 and no suppression event in the last 50 years.

@@

Although the south of Iberian Peninsula suffered severe drought events in the last decades [*e.g* 2005 and 2012 were considered two of the worst recorded drought; @GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of *Q. pyrenaica* (78.9 % of the pixels showed a positive trend during the 2000 to 2016 period). This confirms previous findings that pointed out an increase in primary productivity for these forests [@PerezLuque2015onto], being the most productive ecosystems in this mountain region during that period [@Alcaraz2016obsnev_ndvi]. For tree-growth, we also observed positive trends in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4). Similar long-term trends were described for this species along their distribution range [@GeaIzquierdo2014; @RubioCuadrado2018] and could be related to the rising of the temperatures in the last decades [@GeaIzquierdo2014]. Howewer, the observed pattern for our sites (a rear-edge) differs from the decline trends observed for several Mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P. sylvestris*; @SanchezSalguero2012; @Camarero2015; *Fagus sylvatica*; @Dorado2017].

@@

@@

")
```
```{r echo=FALSE, results='asis', eval=show_text}
cat("
### OTRA COSA

In fact, several authors have raise the need to consider land use y drought de forma conjunta en las environmental y forest management .... (ver Peñuelas et al 2017 y tb Doblas Miranda )

Environmental and forest management policies should take into account all these characteristics of Mediterranean forests and the social, environmental and climatic conditions that are projected for the coming years and decades.
")
```
```{r echo=FALSE, results='asis', eval=show_text}
cat("
#### Climate and tree relationship

@@

- además (Vulnerabilidad de los bosques a los eventos climáticos extremos ...):

@@

### 
Nosotros hemos obtenido una relación entre la severidad de la sequía y la resiliencia mostrada por los robledales en su rear edge. (ver gráfica )

@@

Methodological approach

comentario: tengo que rematar este apartado.

Vegetation reflects the environmental conditions, and the effects of drought on vegetation can be observed using information from several methodological approaches [see @Norman2016 for a review]. Several works revealed the utility of remote-sensing [*e.g*. @Zhang2013; @AghaKouchak2015] and of the dendrochronology [*e.g.* @Eilmann2012; @Bhuyan2017] to quantify the drought impacts on vegetation at different spatial and temporal scales, but to our knowledge the combination of both approaches is scarce [@VicenteSerrano2013; @VicenteSerrano2016; @Bhuyan2017b; @Wu2018]. Estos trabajos han observado que existe de forma general a positive relationship between vegetation indices derived from remote sensing and annual tree growth [@VicenteSerrano2016; @Bhuyan2017b; @Gazol2018].

Aunque existen algunos trabajos que han utilizado RWI y remote sensing nuestra aproximación (creo) es interesante, sobre todo porque la aplicamos al estudio de poblaciones que están en su límite de distribución (rear-edge) [@Jump2010]. Los estudios sobre poblaciones localizadas en el rear-edge de su distribución requieren de aproximaciones multidisciplinares que analicen la respuesta de las poblaciones al cambio climático y su resiliencia utilizando diferentes aproximaciones metodológicas, por ejemplo combinando dendro y remote sensing [@Jump2010].

- Nosotros aqui utilizamos la combinación de técnicas de remote sensing con dendro para analizar la vulnerabilidad de poblaciones de Q. pyrenaica situadas en su reaar edge frente a los eventos de sequía.

- Hemos encontrado diferencias en cuanto a la resiliencia utilizando RS y RWI. Diferente sensibilidad de satélite frente a los datos de campo. Esto también se ha observado en @Gazol2018, que analizaron la resiliencia a la sequia de 1986, 1994-1995, 1999 y 2005, usando RS y TR. Encontraron que los datos de TR son mas sensibles para la resiliencia del bosque a la sequía que los datos de RS

- De hecho, @Jump2010 recomienda el uso de una aprox combinda de dendro, remote sensing y ground-based assessment para analizar los efectos del cambio global en las poblaciones situadas en el rear edge.

- La utilización de datos de tree-ring nos permite contextualizar los resultados obtenidos con remote sensing, ya que los datos de tree-ring reflejan anomalías en el crecimiento de los árboles (inducidas por clima o por disturbances) durante décadas o siglos [@Babst2017] (... Combined analyses may also allow climate-induced variability in forest growth to be disentangled from that driven by community-level ecological processes. )

- Nota from @Jump2010: Whilst lower resolution data, whether from remote sensing or ground-based monitoring, can inform on widespread regional changes in forest condition, these data are often not adequate for monitoring changes occurring at equatorial range margins owing to their complex distribution and/or topographical variability. In such regions, targeted collection of high-resolution data is necessary in order to identify currently occurring changes and predict the magnitude and spatial distribution of future decline. [@Jump2010]

>>>>

During the second half of the twentieth century the Iberian Peninsula suffered several extreme drought episodes (*e.g*. 1994-1995, 1999, 2005, 2012) [@VicenteSerrano2014; @Guerreiro2017].The 2005 and 2012 drought events have been documented among the worst droughts in the last decades for the southern Iberian Peninsula [@GarciaHerrera2007; @Gouveia2014; @Gouveia2015; @Trigo2013; @Guerreiro2017; @Pascoa2017] and they were characterized as extreme drought in our climatic data (Figures S1-S2; Table S3). We focused on these two drought events because they were included in the period where there is availability of high-spatial resolution remote sensing information (MODIS started on 2000; see below). Nevertheless, for radial growth-time series, a greater number of older drought events were also analyzed to contextualize results obtained in 2005 and 2012 drought events and analyze forest resilience to drought in a longer term (see Table S3). The identification of a drought event were done using the Standardized Precipitation-Evapotranspiration Index (SPEI) [@VicenteSerrano2010] following similar procedure as proposed by @Spinoni2015. We used SPEI-12 months scale for all 0.5º grid cells covering Sierra Nevada obtained from Global SPEI Database (http://spei.csic.es/database.html). A severe drought event starts in the month when SPEI falls below the threshold of -1.28 [@Spinoni2017a; @Pascoa2017]. A drought event is considered only when SPEI values are below the threshold for at least two consecutive months [*e.g.* @Spinoni2015; @Spinoni2017a]. Several indicators were computed for each drought event: ***duration*** as the number of consecutive months with the SPEI lower than a certain threshold; ***severity*** of a drought event is the sum of the SPEI values (absolute values) during the duration of the drought event; the ***intensity*** and the ***Lowest SPEI*** refer to the mean and lowest value of SPEI respectively during the drought event duration.

@@

The analysis of growth changes revealed differences in forest history between sites (Figures 5 and S4). Northern site (SJ) showed two release events (GC > 50 %) detected at stand-wise scale (occurring in more than 50 % of sampled trees): the first during the 1940 decade and the second in the period 1995-2000. These periods alternate with periods of supression. Southern sites (CA-High and CA-Low) showed no release events except for CA-High at the beginning of the 1830 and no suppression events in the last 50 years.

@@

Although the south of Iberian Peninsula suffered severe drought events in the last decades [*e.g* 2005 and 2012 were considered two of the worst recorded drought; @GarciaHerrera2007; @Trigo2013; @VicenteSerrano2014] (Table S3), we found a positive trend for vegetation greenness of *Q. pyrenaica* (78.9 % of the pixels showed a positive trend during the 2000 to 2016 period). This confirms previous findings that pointed out an increase in primary productivity for these forests [@PerezLuque2015onto], being the most productive ecosystems in this mountain region during that period [@Alcaraz2016obsnev_ndvi]. For tree-growth, we also observed positive trends in the last decades, particularly for the southern high-elevation site (CA-High, Figure 4). Similar long-term trends were described for this species along their distribution range [@GeaIzquierdo2014; @RubioCuadrado2018] and could be related to the rising of the temperatures in the last decades [@GeaIzquierdo2014] particularly after 1970 [@BladeCastroDiez2010CeEppyf; @VicenteSerranoetal2017CIG]. Howewer, the observed pattern for our sites (a rear-edge) differs from the decline trends observed for several Mediterranean and temperate tree-species located in their rear-edges [*e.g.*, *Pinus nigra*; *P. sylvestris*; @SanchezSalguero2012; @Camarero2015; *Fagus sylvatica*; @Dorado2017].

@@

Greenness and tree-growth of *Q. pyrenaica* showed different sensitivity to severe droughts. The exploration of the average reduction for the 2005 and 2012 droguht events
revaled that EVI was reduced to 90.25 % (Table S1), whereas BAI was reduced to 70.83 % (Table S2). This suggests a lower sensitivity of the greenness than tree-growth to drought, particularly for sites under drier climate. For instance, during 2005 drought event, the BAI were reduced to 44.5 % respect to that of the preceding period in northern site (SJ), whereas the EVI was reduced to 81.9 % for the same site during the same drought event (Tables S1 and S2).

The different sensitivity

Causas: 
Gazol
EVI of the understory (Dioniosio) 
Different scales

@@

")
```
```{r echo=FALSE, results='asis', eval=show_text}
cat("
#### Climate and tree relationship

@@

@@

###

@@
